---
title: "DS311 - R Lab Assignment"
author: "Shaun Kumar"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    theme: united
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Assignment 1

* In this assignment, we are going to apply some of the build in data set in R for descriptive statistics analysis.
* To earn full grade in this assignment, students need to complete the coding tasks for each question to get the result.
* After finished all the questions, knit the document into HTML format for submission.

### Question 1

Using the **mtcars** data set in R, please answer the following questions.

```{r}
# Loading the data
data(mtcars)

# Head of the data set
head(mtcars)
```

a. Report the number of variables and observations in the data set.

```{r}
# Enter your code here!
num_obs <- nrow(mtcars)
num_vars <- ncol(mtcars)

# Answer:
print(paste("There are total of",num_vars, "variables and",num_obs, "observations in this data set."))
```



b. Print the summary statistics of the data set and report how many discrete and continuous variables are in the data set.

```{r}
# Enter your code here!
summary(mtcars)

discrete_vars <- sapply(mtcars, function(x) is.integer(x) || (is.numeric(x) && length(unique(x)) < 10))
continuous_vars <- !discrete_vars

num_discrete <- sum(discrete_vars)
num_continuous <- sum(continuous_vars)

# Answer:
print(paste("There are",num_discrete, "discrete variables and",num_continuous, "continuous variables in this data set."))
```



c. Calculate the mean, variance, and standard deviation for the variable **mpg** and assign them into variable names m, v, and s. Report the results in the print statement.

```{r}
# Enter your code here!
m <- round(mean(mtcars$mpg),2)
v <- round(var(mtcars$mpg),2)
s <- round(sd(mtcars$mpg),2)


 print(paste("The average of Mile Per Gallon from this data set is", m , "with variance", v , "and standard deviation", s, "."))
```



d. Create two tables to summarize 1) average mpg for each cylinder class and 2) the standard deviation of mpg for each gear class.

```{r}
# Enter your code here!
avg_mpg_by_cyl <- aggregate(mpg ~ cyl, data = mtcars, FUN = mean)
sd_mpg_by_gear <- aggregate(mpg ~ gear, data = mtcars, FUN = sd)

print(avg_mpg_by_cyl)
print(sd_mpg_by_gear)

```



e. Create a crosstab that shows the number of observations belong to each cylinder and gear class combinations. The table should show how many observations given the car has 4 cylinders with 3 gears, 4 cylinders with 4 gears, etc. Report which combination is recorded in this data set and how many observations for this type of car.

```{r}
# Enter your code here!
crosstab <- table(mtcars$cyl, mtcars$gear)
max_count <- max(crosstab)
max_position <- which(crosstab == max_count, arr.ind = TRUE)

most_common_cyl <- as.numeric(rownames(crosstab)[max_position[1]])
most_common_gear <- as.numeric(colnames(crosstab)[max_position[2]])

print(paste("The most common car type in this data set is car with",most_common_cyl,"cylinders and",most_common_gear,"gears. There are total of",max_count,"cars belong to this specification in the data set."))
```


*** 

### Question 2

Use different visualization tools to summarize the data sets in this question.

a. Using the **PlantGrowth** data set, visualize and compare the weight of the plant in the three separated group. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your findings.

```{r}
# Load the data set
data("PlantGrowth")

# Head of the data set
head(PlantGrowth)

# Enter your code here!
# Set a CRAN mirror before installing packages
options(repos = c(CRAN = "https://cran.rstudio.com"))
install.packages("ggplot2")
library(ggplot2)

ggplot(PlantGrowth, aes(x = group, y = weight, fill = group)) +
  geom_boxplot() +
  labs(
    title = "Comparison of Plant Weights Across Groups",
    x = "Treatment Group",
    y = "Plant Weight"
  ) +
  theme_minimal() +
  theme(legend.position = "none") 

```

Result:

=> Report a paragraph to summarize your findings from the plot!
The boxplot shows how plant weights differ between the three groups: control, treatment 1 (trt1), and treatment 2 (trt2). The control group has a fairly steady range of plant weights, with most plants weighing around 5. Treatment 1 seems to result in slightly lower plant weights, while treatment 2 shows slightly higher weights compared to the control group. However, the ranges of weights for all three groups overlap, so it’s not clear if the differences are meaningful just by looking at the plot.


b. Using the **mtcars** data set, plot the histogram for the column **mpg** with 10 breaks. Give labels to the title, x-axis, and y-axis on the graph. Report the most observed mpg class from the data set.

```{r}
hist_result <- hist(
  mtcars$mpg,
  breaks = 10,
  main = "Distribution of Miles Per Gallon (mpg)",
  xlab = "Miles Per Gallon (mpg)",
  ylab = "Frequency",
  col = "lightblue",
  border = "black"
)

max_bin_index <- which.max(hist_result$counts)
most_common_range <- paste(round(hist_result$breaks[max_bin_index], 1), 
                           "-", 
                           round(hist_result$breaks[max_bin_index + 1], 1))


print(paste("Most of the cars in this data set are in the class of",most_common_range,"miles per gallon."))
```



c. Using the **USArrests** data set, create a pairs plot to display the correlations between the variables in the data set. Plot the scatter plot with **Murder** and **Assault**. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your results from both plots.

```{r}
# Load the data set
data("USArrests")

# Head of the data set
head(USArrests)

# Enter your code here!
pairs(USArrests, main = "Pairs Plot of USArrests Data")

plot(USArrests$Murder, USArrests$Assault,
     main = "Scatter Plot of Murder vs Assault",
     xlab = "Murder Rate",
     ylab = "Assault Rate",
     col = "blue",
     pch = 19)  # pch=19 gives solid circles

```

Result:

=> Report a paragraph to summarize your findings from the plot!

The plot shows the relationship between the different crime rates in the USArrests dataset. It highlights a clear positive correlation between Murder and Assault, where states with higher murder rates tend to have higher assault rates as well. The scatter plot further confirms this, showing an upward trend between these two variables.  Overall, the plots suggest that murder and assault are closely related.


***

### Question 3

Download the housing data set from www.jaredlander.com and find out what explains the housing prices in New York City.

Note: Check your working directory to make sure that you can download the data into the data folder. 

```{r, echo=FALSE}
# Load and clean the housing data set 
download.file(url = 'https://www.jaredlander.com/data/housing.csv',
              destfile = '../data/housing.csv', mode = 'wb')
housingData <- read.csv('../data/housing.csv')
housingData <- subset(housingData, 
                      select = c("Neighborhood", "Market.Value.per.SqFt", "Boro", "Year.Built"))
housingData <- na.omit(housingData)                   
```


a. Create your own descriptive statistics and aggregation tables to summarize the data set and find any meaningful results between different variables in the data set.

```{r}
# Head of the cleaned data set
head(housingData)

# Enter your code here!
summary(housingData)
aggregate(Market.Value.per.SqFt ~ Boro, data = housingData, FUN = mean)
aggregate(Year.Built ~ Boro, data = housingData, FUN = mean)
cor(housingData$Market.Value.per.SqFt, housingData$Year.Built)
table(housingData$Boro, housingData$Neighborhood)

install.packages("dplyr")

library(dplyr)

housingData %>%
  group_by(Boro) %>%
  summarise(
    Avg_Value = mean(Market.Value.per.SqFt),
    Avg_Year = mean(Year.Built),
    Count = n()
  )
```


b. Create multiple plots to demonstrates the correlations between different variables. Remember to label all axes and give title to each graph.

```{r}
# This shows if newer houses are generally more expensive per square foot
plot(housingData$Year.Built, housingData$Market.Value.per.SqFt,
     main = "Market Value vs. Year Built",
     xlab = "Year Built",
     ylab = "Market Value per SqFt",
     pch = 19,        # solid circles
     col = "blue")    # color of points

boxplot(Market.Value.per.SqFt ~ Boro, data = housingData,
        main = "Market Value per SqFt by Borough",
        xlab = "Borough",
        ylab = "Market Value per SqFt",
        col = "lightgreen")

# This shows how prices are distributed 
hist(housingData$Market.Value.per.SqFt,
     main = "Distribution of Market Value per SqFt",
     xlab = "Market Value per SqFt",
     ylab = "Frequency",
     col = "orange",
     breaks = 30)   # number of bins

# This graoh shows which boroughs are most/least expensive on average
library(dplyr)

boro_avg <- housingData %>%
  group_by(Boro) %>%
  summarise(Avg_Value = mean(Market.Value.per.SqFt))

barplot(boro_avg$Avg_Value,
        names.arg = boro_avg$Boro,
        main = "Average Market Value per SqFt by Borough",
        xlab = "Borough",
        ylab = "Average Market Value per SqFt",
        col = "lightblue")

```

c. Write a summary about your findings from this exercise.

=> Enter your answer here!

In this exercise, we analyzed the housing dataset containing information on neighborhood, borough, market value per square foot, and year built. After cleaning the data, descriptive statistics and graphical visualizations demonstrated that location plays a more critical role in determining property value than building age. While newer constructions often had higher market values, older properties in desirable locations remained competitively priced. The distribution of market values was heavily right-skewed, with a concentration of moderately priced properties and a smaller number of high-value outliers. Significant variability within boroughs further highlighted the diverse nature of local housing markets.

